---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(testwhat)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.checker = testwhat_learnr)
```

## Topic 1

### Exercise 

*Here is a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

```{r two-plus-two-solution}
2 + 2
```

```{r two-plus-two-check}
ex() %>% check_operator("+") %>% check_result() %>% check_equal()
```

### Exercise with Code

*Here is an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

```{r add-function-solution}
# Correct function form
add <- function(x, y) {
  return(x + y)
}

# Function call with arbitrary numbers
add(6, 1)
```

```{r add-function-check}
# Checks function and calls with various inputs
ex() %>% check_fun_def("add") %>% {
  check_arguments(.)
  check_call(., 1, 2) %>% check_result() %>% check_equal()
  check_call(., -1, 2) %>% check_result() %>% check_equal()
  check_call(., 1, -2) %>% check_result() %>% check_equal()
  check_call(., -1, -2) %>% check_result() %>% check_equal()
  check_body(.) %>% {
    check_operator(., "+")
  }
}

# Checks the function is called correctly
ex() %>% check_function("add")
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

```{r print-limit-solution}
# n argument limits the number of lines, default is 6L
head(mtcars, n=5L)
```

```{r print-limit-check}
# Checks that head is called and both arguments are correct
ex() %>% check_function("head") %>% {
  check_arg(., "x") %>% check_equal()
  check_arg(., "n") %>% check_equal()
}
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Here some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

